---
title: "`r config$publication_0404_name`"
pagetitle: MUMH England quarterly summary
always_allow_html: true
output:
  word_document:
    toc: yes
    toc_depth: '3'
  html_document:
    anchor_sections: no
    css: www/style.css
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: no
---

<script>
  $(document).ready(function() {
    $('#header').prepend('<img src=`r knitr::image_uri("www/bsa_logo.svg")` alt=\"NHSBSA logo\" style=\"position:absolute; top: 0; right: 0; padding: 10px; width: 25%;\">');
  });
</script>

<style type="text/css">  body, td {
    font-size: 16px;
    font-family: sans-serif;
  }
  .list-group-item.active {
    background-color: #005EB8!important;
    border-color: #005EB8!important;
  }
    a {
        color: #005EB8!important;
    }
    .tocify {
        border-radius: 0px!important;
    }
</style>
  <html lang="en">

```{r setup, include=FALSE}
# set code chunk options to disable echo by default
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      message = FALSE)

```

<main>

# `r config$publication_sub_name` {.toc-ignore}

Published `r config$publication_date` 

## Summary

Between July and September 2023:

::::{.row style="display: flex; padding-bottom: 15px;"}
:::{.col-md-6}
`r infoBox_no_border("", text = "<b>There were 720,000 CNS stimulants and drugs for ADHD items prescribed to an estimated 210,000 identified patients.</b>", width = "100%")`
:::
:::{.col-md-6}
`r infoBox_no_border("", text = "<b>The cost of prescribed CNS stimulants and drugs for ADHD items was £31 million.</b>", width = "100%")`
:::
::::

---

## 1. Things you should know

`r infoBox_border(" ", text = "CNS stimulants and drugs used for ADHD are described in the BNF 68 section 4.4: 

* BNF chemical substance 0404000D0 - Caffeine
* BNF chemical substance 0404000E0 - Caffeine citrate
* BNF chemical substance 0404000L0 - Dexamfetamine sulfate
* BNF chemical substance 0404000M0 - Methylphenidate hydrochloride
* BNF chemical substance 0404000R0 - Modafinil
* BNF chemical substance 0404000S0 - Atomoxetine hydrochloride
* BNF chemical substance 0404000T0 - Dexmethylphenidate hydrochloride
* BNF chemical substance 0404000U0 - Lisdexamfetamine dimesylate
* BNF chemical substance 0404000V0 - Guanfacine
* BNF chemical substance 0404000W0 - Pitolisant hydrochloride", width = "100%")`

### 1.1. Central nervous system (CNS) stimulants and drugs used for ADHD {.toc-ignore}

Attention deficit hyperactivity disorder (ADHD) and attention deficit disorder (ADD) refer to a range of problem behaviours associated with poor attention span. These may include impulsiveness and hyperactivity, as well as inattentiveness; behaviours that often prevent children and adults from learning and socialising. ADHD is sometimes referred to as hyperkinetic disorder (HD).

You can find more information about [ADHD](https://www.nhs.uk/conditions/attention-deficit-hyperactivity-disorder-adhd/) on the NHS website.

---

## 2. Results - Central nervous system (CNS) stimulants and drugs used for ADHD

#### Figure 1: Number of prescribed items and identified patients by financial quarter - BNF 4.4: Central nervous system (CNS) stimulants and drugs used for ADHD

```{r quarterly_0404, out.width="100%"}

figure_1_0404
  
```
`r get_download_button(title = "Download chart data", data = figure_1_data_0404, filename = "figure_1_0404")`

::::{.row style="display: flex; padding-bottom: 15px;"}
:::{.col-md-4}
`r infoBox_no_border("", text = "<b>720,000 CNS stimulants and drugs for ADHD items prescribed in Q2 2023/24.</b>", width = "100%")`
:::
:::{.col-md-4}
`r infoBox_no_border("", text = "<b>210,000 identified patients received prescribing for CNS stimulants and drugs for ADHD section in Q2 2023/24.</b>", width = "100%")`
:::
:::{.col-md-4}
`r infoBox_no_border("", text = "<b>There was a 0.7% increase in items and a 2.6% increase in patients on the previous quarter. Items and patients continue to trend upwards for CNS stimulants and drugs for ADHD items. </b>", width = "100%")`
:::
::::


#### Figure 2: Cost of prescribing by financial quarter - BNF 4.4: Central nervous system (CNS) stimulants and drugs used for ADHD

```{r quarterly_0404_nic, out.width="100%"}

figure_2_0404
  
```
`r get_download_button(title = "Download chart data", data = figure_2_data_0404, filename = "figure_2_0404")`

::::{.row style="display: flex; padding-bottom: 15px;"}
:::{.col-md-6}
`r infoBox_no_border("", text = "<b>In Q2 2023/24 the cost of prescribed CNS stimulants and drugs for ADHD items was £31 million. </b>", width = "100%")`
:::
:::{.col-md-6}
`r infoBox_no_border("", text = "<b>This was a 4.4% increase on the previous quarter, so costs increased by a higher proportion than items.</b>", width = "100%")`
:::
::::

#### Figure 3: Number of prescribed items and identified patients by month - BNF 4.4: Central nervous system (CNS) stimulants and drugs used for ADHD

```{r monthly_0404, out.width="100%"}

figure_3_0404
```
`r get_download_button(title = "Download chart data", data = figure_3_data_0404, filename = "figure_3_0404")`

::::{.row style="display: flex; padding-bottom: 15px;"}
:::{.col-md-6}
`r infoBox_no_border("", text = "<b>CNS stimulants and drugs for ADHD items increased by 19% over the last 12 months, compared to the previous 12 month period. </b>", width = "100%")`
:::
:::{.col-md-6}
`r infoBox_no_border("", text = "<b>Over the last 12 months, an average of 130,000 patients per month received prescribing for a CNS stimulants and drugs for ADHD item.</b>", width = "100%")`
:::
::::

---

## 3. Sections in this release

* Medicines used in Mental Health [summary](`r config$html_top_link`)
* BNF Section 4.3 - [Antidepressant drugs](`r config$html_0403_link`)
* BNF Section 4.1 - [Hypnotics and anxiolytics](`r config$html_0401_link`)
* BNF Section 4.2 - [Drugs used in psychoses and related disorders](`r config$html_0402_link`)
* BNF Section 4.11 - [Drugs for dementia](`r config$html_0411_link`)
* [Prescribing during the COVID-19 pandemic](`r config$html_model_link`)
* Medicines used in Mental Health [background](`r config$background_link`)