---
title: "Medicines Used in <br> Mental Health - Antibiotics"
pagetitle: MUMH England quarterly summary
always_allow_html: true
output:
  word_document:
    toc: yes
    toc_depth: '3'
  html_document:
    anchor_sections: no
    css: www/style.css
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: no
---

<script>
  $(document).ready(function() {
    $('#header').prepend('<img src=`r knitr::image_uri("www/bsa_logo.svg")` alt=\"NHSBSA logo\" style=\"position:absolute; top: 0; right: 0; padding: 10px; width: 25%;\">');
  });
</script>

<style type="text/css">  body, td {
    font-size: 16px;
    font-family: sans-serif;
  }
  .list-group-item.active {
    background-color: #005EB8!important;
    border-color: #005EB8!important;
  }
    a {
        color: #005EB8!important;
    }
    .tocify {
        border-radius: 0px!important;
    }
</style>
  <html lang="en">

```{r setup, include=FALSE}
# set code chunk options to disable echo by default
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      message = FALSE)

```

<main>

# England April 2017 to June 2023 {.toc-ignore}

Published 14 September 2023

`r mumhquarterly::infoBox_border(header = "Changes to these statistics", text = paste0("For this release, we have split the publication into smaller summaries for each BNF Section. We are interested in any feedback about the publication or these changes, which you can send by using our ", htmltools::HTML("<a href='https://online1.snapsurveys.com/Official_Statistics_Feedback'>Official Statistics feedback survey</a>"),"."), width = "100%")`

## Summary

Between April and June 2023:

::::{.row style="display: flex; padding-bottom: 15px;"}
:::{.col-md-4}
`r mumhquarterly::infoBox_no_border("", text = "<b>22 million antidepressant items were prescribed to an estimated 6.7 million identified patients.</b>", width = "100%")`
:::
:::{.col-md-4}
`r mumhquarterly::infoBox_no_border("", text = "<b>The costs of prescribed antidepressant items was £59.8 million..</b>", width = "100%")`
:::
:::{.col-md-4}
`r mumhquarterly::infoBox_no_border("", text = "<b>Prescribing of drugs for dementia items remained below expected values based on pre-pandemic trends.</b>", width = "100%")`
:::
::::




---

## 1. Things you should know
### 1.1. Scope {.toc-ignore} 

`r mumhquarterly::infoBox_border("Background ", text = "The [NHS Long Term Plan](https://www.longtermplan.nhs.uk/) published in 2019 focuses on improving mental health care in the UK. It aims to provide more funding for mental health services, particularly for children and people with dementia. The [Core20PLUS5](https://www.england.nhs.uk/about/equality/equality-hub/core20plus5/) approach  to reduce health inequalities also seeks to improve community care for those with severe mental illnesses.", width = "100%")`

This publication provides information on medicines used to treat depression in England. It covers medicines prescribed in England that are then dispensed in the community in England, Scotland, Wales, Isle of Man of the Channel Islands. The publication does not include data on medicines used in hospitals, prisons or prescribed by private doctors. It also explores how the COVID-19 pandemic has affected the use of these medicines.

Five summaries total are available covering medicines used in mental health, the others being for [anxiety](https://www.dummylink.com), [psychosis](https://www.dummylink.com), [attention deficit hyperactivity disorder (ADHD)](https://www.dummylink.com), and [dementia](https://www.dummylink.com).

These summary narratives do not discuss the demographic data associated with the provision of these medicines, but this information is available in the supporting summary tables which accompany this release. Medicines are not the only way to treat mental health conditions, and this publication does not capture information on the use of other mental health services, such as psychological therapies. Therefore, this publication may not give a complete picture of the number of people receiving treatment for mental health conditions. Information on where to find related statistics can be found in Section 4.

### 1.2. Definitions {.toc-ignore}

`r mumhquarterly::infoBox_border("Item", text = "A single unit of medication listed separately on a prescription form. In this publication, an example of an item would be Fluoxetine 20mg tables x56.", width = "100%")`

`r mumhquarterly::infoBox_border("Patient", text = "A unique NHS number verified from a prescription form. Verification of an NHS number for patient identification is done by the NHS Personal Demographics Service (PDS).", width = "100%")`

`r mumhquarterly::infoBox_border("NIC", text = "The Net Ingredient Cost (NIC) is the basic price of the medication and the quantity prescribed. It does not include other fees incurred by dispensing contractors, such as controlled drug fees or the single activity fee. The basic price is determined by the [Drug Tariff](https://www.nhsbsa.nhs.uk/pharmacies-gp-practices-and-appliance-contractors/drug-tariff) or by the manufacturer, wholesaler, or supplier of the product.", width = "100%")`

`r mumhquarterly::infoBox_border("Classification", text = "This publication uses the British National Formulary (BNF), which lists medicines used in the UK and classifies them according to their primary therapeutic use. Medication may also be prescribed for other purposes beyond their intended use. For example, certain antidepressants may be recommended to individuals experiencing chronic pain. The NHSBSA does not capture the clinical indication of a prescription during processing.", width = "100%")`

### 1.3. Time periods {.toc-ignore}

The data in this publication covers the period from April 2017 to June 2023. Q1, Q2, Q3, and Q4 refer to different periods within a financial year. You can find more information on the [Medicines Used in Mental Health - England webpage](https://www.nhsbsa.nhs.uk/statistical-collections/medicines-used-mental-health-england).

These statistics cover the period of the coronavirus (COVID-19) pandemic. A detailed [timeline](https://commonslibrary.parliament.uk/research-briefings/cbp-9068/) of events and restrictions in the UK during this time can be found on the UK parliament website. The impact of COVID-19 should be considered when comparing across time periods 

### 1.4. Patient identification {.toc-ignore}

When the NHSBSA processes prescriptions it is not always possible to capture the NHS number of the patient. Table 1 shows the proportion of items for which a patient could be identified. This means that the data relating to patient counts represents most, but not all, patients.

Due to an increase in digital prescription processing through the Electronic Prescription Service (EPS) during the COVID-19 pandemic, more patients were identified in 2020/21 and 2021/22 compared to previous years. As patient identification rates increased, any increases in the number of identified patients between periods are likely to be an overestimate of the actual increase in patient numbers. This is because the proportion of patients who could be identified has increased. Conversely, any decrease over the same period is likely to be an underestimate of the actual decrease.

These statistics do not include any information that is personally identifiable. You can find more information about how the NHSBSA protect personal information in the [confidentiality and access statement](https://www.nhsbsa.nhs.uk/policies-and-procedures). 

#### Table 1: The proportion of items for which an NHS number was recorded for BNF 4.3: Antidepressant drugs for previous 4 quarters
```{r pat_ident_rates, out.width="100%"}

table_1_0403

```
`r get_download_button(title = "Download data table", data = table_1_data_0403, filename = "table_1_0403")`

---

## 2. Results and commentary - Antidepressants

`r mumhquarterly::infoBox_border(" ", text = "Antidepressant drugs are effective for treating moderate to severe depression, but are not usually recommended for the routine treatment of mild depression. In those cases, psychological therapy should be considered initially. However, a trial of antidepressant therapy may be considered in cases resistant to psychological treatments or associated with other problems. The National Institute for Health and Care Excellence (NICE) has [released updated guidance](https://www.nice.org.uk/news/nice-draft-quality-standard-depression-adults-update-2023) to recommend adult patients stopping antidepressants should have their dose reduced in stages across time.

Antidepressant drugs are described in the BNF 68 section 4.3:

* BNF paragraph 4.3.1 - Tricyclic and related antidepressant drugs
* BNF paragraph 4.3.2 - Monoamine-oxidase inhibitors (MAOIs)
* BNF paragraph 4.3.3 - Selective serotonin re-uptake inhibitors (SSRIs)
* BNF paragraph 4.3.4 - Other antidepressant drugs

More information about these medicines can be found in section 4 of this summary.", width = "100%")`

#### Figure 1: Number of prescribed items and identified patients by financial quarter - BNF 4.3: Antidepressant drugs

```{r quarterly_003, out.width="100%"}

figure_1
  
```
`r get_download_button(title = "Download chart data", data = figure_1_data, filename = "figure_1")`

::::{.row style="display: flex; padding-bottom: 15px;"}
:::{.col-md-4}
`r mumhquarterly::infoBox_no_border("", text = "<b>22 million antidepressant items prescribed in Q1 2023/24.</b>", width = "100%")`
:::
:::{.col-md-4}
`r mumhquarterly::infoBox_no_border("", text = "<b>6.7 million identified patients received an antidepressant in Q1 2023/24.</b>", width = "100%")`
:::
:::{.col-md-4}
`r mumhquarterly::infoBox_no_border("", text = "<b>Items and patients continue to trend upwards for antidepressants.</b>", width = "100%")`
:::
::::

In April to June 2023, prescribing of antidepressant drug items increased to 21.9 million items, 6.58% more than the same time last year. This continues the long-term trend for antidepressant prescribing, which has risen by 25% since 2017/18. 

An estimated 6.71 million identified patients were prescribed at least one antidepressant item between April to June 2023. This was not a notable change from the previous 3 months, but estimated identified patients are increasing along the same long-term trend as items.   


#### Figure 2: Cost of prescribing by financial quarter - BNF 4.3: Antidepressant drugs

```{r quarterly_003_nic, out.width="100%"}

figure_2
  
```
`r get_download_button(title = "Download chart data", data = figure_2_data, filename = "figure_2")`

In April to June 2023, the cost of prescribing of antidepressant drug items decreased to £59.8 million, which was 1.05% less than the previous quarter. The long-term trend for the cost of antidepressant prescribing, has varied more than volume due to price fluctuations. 

#### Figure 3: Number of prescribed items and identified patients by month - BNF 4.3: Antidepressant drugs

```{r monthly_0403, out.width="100%"}

figure_3
```
`r get_download_button(title = "Download chart data", data = figure_3_data, filename = "figure_3")`

::::{.row style="display: flex; padding-bottom: 15px;"}
:::{.col-md-6}
`r mumhquarterly::infoBox_no_border("", text = "<b>On average, 4.6 million patients received an antidepressant each month in 2022.</b>", width = "100%")`
:::
:::{.col-md-6}
`r mumhquarterly::infoBox_no_border("", text = "<b>Antidepressant items increased by 3.2% between 2021 and 2022.</b>", width = "100%")`
:::
::::


Figure 3 shows antidepressant prescribing continues to have regular and seasonal patterns. For all BNF sections in this publication, months with fewer dispensing days, such as February, tend to have less prescribing. Despite the month-to-month variation, antidepressants show a clear increase over time.

In in the 12 months July 2022 to June 2023, there was a monthly average of 4.64 million identified patients receiving at least one antidepressant item. This was a small increase from the monthly average of 4.53 million identified patients in the previous 12 months.

There were 86.9 million antidepressant drug items prescribed in the 12 months July 2022 to June 2023, a 3.22% increase from 84.2 million items in July 2021 to June 2022. 

---

## 3. Background

This publication provides information on medicines used to treat mental health conditions in England. It covers medicines prescribed in England that are then dispensed in the community in England, Scotland, Wales, Isle of Man of the Channel Islands by a pharmacy, appliance contractor, dispensing doctor, or have been personally administered by a GP practice . The publication does not include data on medicines used in hospitals, prisons or prescribed by private doctors.

### 3.1. Antidepressant drugs {.toc-ignore}

Antidepressant drugs are licensed to treat major depression. Health professionals use the words depression, depressive illness or clinical depression to refer to depression. It is a serious illness and very different from the common experience of feeling unhappy or fed up for a short period of time. Depressed people may have feelings of extreme sadness that can last for a long time. These feelings are severe enough to interfere with daily life, and can last for weeks, months or years, rather than days.

It should be noted that antidepressant drugs are used for indications other than depression, for example migraine, chronic pain, myalgic encephalomyelitis (ME), or a range of other conditions. Clinical indication isn't captured by the NHSBSA. Therefore, the statistics on these drugs do not relate solely to prescribing for depression.

You can find more information about [depression](https://www.nhs.uk/conditions/clinical-depression/) on the NHS website.

### 3.2. Related statistics and resources {.toc-ignore}

Prescribing of medications in primary care is only part of the big picture for mental health care in England. The [NHS Digital mental health data hub](https://digital.nhs.uk/data-and-information/data-tools-and-services/data-services/mental-health-data-hub) has links to statistics on a range of related topics which may provide further context. These include a mental health bulletin, reports on children and young people's mental health surveys, dementia assessment and referral statistics, and more.

---

## 4. About these statistics

This publication is part of a series by the NHSBSA. This quarterly summary is intended to supplement the [annual summary statistics](https://www.nhsbsa.nhs.uk/statistical-collections/medicines-used-mental-health-england/medicines-used-mental-health-england-201516-202122) released in each year and previous quarterly summaries. To learn more about how this series compares to our other publications, you can view the Official Statistics guidance table on our [statistical collections page](https://www.nhsbsa.nhs.uk/statistical-collections). 

### 4.1. Changes to this publication {.toc-ignore}

For this release we have split the publication into five smaller summaries, one for each BNF section.

### 4.2. Patient counts {.toc-ignore}

The patient counts shown in these statistics should only be analysed at the level at which they are presented. Adding together any patient counts is likely to result in an overestimate of the number of patients. A patient will be included, or counted, in each category or time period in which they received relevant prescriptions. For example, if a patient received a prescription item for an antidepressant drug in Q1 2020/21 and another in Q2 2020/21, then adding together those totals would count that patient twice. For the same reason, data on patient counts for different BNF sections should not be added together.

### 4.3. Rounding {.toc-ignore}

The high-level figures in this statistical summary have been rounded where appropriate for clarity. In most cases rounding has been to three significant figures in the main text and two significant figures in the key findings and information boxes. This is to make this narrative as accessible as possible to all readers. The summary tables released with this publication allow users to investigate this data at lower levels of granularity. Figures in the supplementary tables have not been rounded.


### 4.4. Averages {.toc-ignore}

Where this document refers to 'average', this is the mean unless otherwise stated. This is calculated by adding the number of items/patients/months together and dividing this by the number of items/patients/months.

### 4.5. Planned changes to this publication {.toc-ignore}

This is an experimental official statistic release. Experimental statistics are newly developed or innovative statistics. These are published so that users and stakeholders can be involved in the assessment of their suitability and quality at an early stage. We will regularly be reviewing the methodology used within the statistics.

---

## 5. Accessibility

If you need information on this website in a different format like accessible PDF, large print, easy read, audio recording or braille, you can contact us by:

**Email**: nhsbsa.accessibility@nhs.net

**Phone**: 0191 203 5318

[Find out about call charges](https://www.nhsbsa.nhs.uk/contact-us/call-charges-and-phone-numbers)

We’ll consider your request and get back to you in 5 working days.

These contact details are only for accessibility queries. This inbox is not for technical queries or IT problems. If you have a query that is not about accessibility, go to the ‘Contact us’ section of this page.

View our [Accessibility statement for Official Statistics Narratives](https://www.nhsbsa.nhs.uk/accessibility-statement-official-statistics-narratives).

---

## 6. Feedback and contact us

Feedback is important to us; we welcome any questions and comments relating to these statistics. 

You can complete a [short survey about this publication](https://online1.snapsurveys.com/interview/caf0d685-0500-48d2-8b96-1c434a956819) to help us improve the Official Statistics that we produce. All responses will remain anonymous and individuals will not be identifiable in any report that we produce.

You can view our [privacy policy](https://www.nhsbsa.nhs.uk/our-policies/privacy) on our website to see how your data is used and stored.

You can contact us by:

**Email:** nhsbsa.statistics@nhs.net

**You can also write to us at:**

NHSBSA - Statistics  
NHS Business Services Authority  
Stella House  
Goldcrest Way  
Newburn Riverside  
Newcastle upon Tyne  
NE15 8NY

**Responsible statistician:** Kirsty Gray

</main>
