---
title: "`r config$publication_0402_name`"
pagetitle: MUMH England quarterly summary
always_allow_html: true
output:
  word_document:
    toc: yes
    toc_depth: '3'
  html_document:
    anchor_sections: no
    css: www/style.css
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: no
---

<script>
  $(document).ready(function() {
    $('#header').prepend('<img src=`r knitr::image_uri("www/bsa_logo.svg")` alt=\"NHSBSA logo\" style=\"position:absolute; top: 0; right: 0; padding: 10px; width: 25%;\">');
  });
</script>

<style type="text/css">  body, td {
    font-size: 16px;
    font-family: sans-serif;
  }
  .list-group-item.active {
    background-color: #005EB8!important;
    border-color: #005EB8!important;
  }
    a {
        color: #005EB8!important;
    }
    .tocify {
        border-radius: 0px!important;
    }
</style>
  <html lang="en">

```{r setup, include=FALSE}
# set code chunk options to disable echo by default
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      message = FALSE)

```

<main>

# `r config$publication_sub_name` {.toc-ignore}

Published `r config$publication_date` 


## Summary

Between July and September 2023:

::::{.row style="display: flex; padding-bottom: 15px;"}
:::{.col-md-6}
`r infoBox_no_border("", text = "<b>3.4 million antipsychotic items were prescribed to an estimated 650,000 identified patients.</b>", width = "100%")`
:::
:::{.col-md-6}
`r infoBox_no_border("", text = "<b>The cost of prescribed antipsychotic items was £39 million.</b>", width = "100%")`
:::
::::


---

## 1. Things you should know

`r infoBox_border(" ", text = "These drugs are used to treat psychoses and related disorders, and are referred to under the umbrella term antipsychotics. These drugs are classified and described under the BNF 68 section 4.2:

* BNF paragraph 4.2.1 - Antipsychotic drugs
* BNF paragraph 4.2.2 - Antipsychotic depot injections
* BNF paragraph 4.2.3 - Drugs for mania and hypomania", width = "100%")`

### 1.1. Drugs used in psychoses and related disorders {.toc-ignore}

People experiencing psychoses may report hallucinations (seeing or hearing things that other people cannot see or hear) or delusional beliefs (believing things that are not actually true), and may exhibit personality changes and thought disorder. As well as this, they may experience difficulties with social interaction and impairment in carrying out daily life activities.

You can find more information about [psychoses and related disorders](https://www.nhs.uk/conditions/psychosis/) on the NHS website.


---

## 2. Results - Drugs used in psychoses and related disorders

#### Figure 1: Number of prescribed items and identified patients by financial quarter - BNF 4.2: Drugs used in psychoses and related disorders

```{r quarterly_0402, out.width="100%"}

figure_1_0402
  
```
`r get_download_button(title = "Download chart data", data = figure_1_data_0402, filename = "figure_1_0402")`

::::{.row style="display: flex; padding-bottom: 15px;"}
:::{.col-md-4}
`r infoBox_no_border("", text = "<b>3.4 million antipsychotic items prescribed in Q2 2023/24.</b>", width = "100%")`
:::
:::{.col-md-4}
`r infoBox_no_border("", text = "<b>650,000 identified patients received an antipsychotic item in Q2 2023/24.</b>", width = "100%")`
:::
:::{.col-md-4}
`r infoBox_no_border("", text = "<b>Antipsychotics continue to show an upwards trend for items and identified patients.</b>", width = "100%")`
:::
::::


#### Figure 2: Cost of prescribing by financial quarter - BNF 4.2: Drugs used in psychoses and related disorders

```{r quarterly_0402_nic, out.width="100%"}

figure_2_0402
  
```
`r get_download_button(title = "Download chart data", data = figure_2_data_0402, filename = "figure_2_0402")`

::::{.row style="display: flex; padding-bottom: 15px;"}
:::{.col-md-6}
`r infoBox_no_border("", text = "<b>The cost of prescribed antipsychotic items was £39 million in Q2 2023/24.</b>", width = "100%")`
:::
:::{.col-md-6}
`r infoBox_no_border("", text = "<b>Since the previous quarter, the cost of antipsychotic prescribing has decreased by 4.3%.</b>", width = "100%")`
:::
::::

#### Figure 3: Number of prescribed items and identified patients by month - BNF 4.2: Drugs used in psychoses and related disorders

```{r monthly_0402, out.width="100%"}

figure_3_0402
```
`r get_download_button(title = "Download chart data", data = figure_3_data_0402, filename = "figure_3_0402")`

::::{.row style="display: flex; padding-bottom: 15px;"}
:::{.col-md-6}
`r infoBox_no_border("", text = "<b>An average of 500,000 patients per month received prescribing for antipsychotic over the last 12 months.</b>", width = "100%")`
:::
:::{.col-md-6}
`r infoBox_no_border("", text = "<b>Antipsychotic items increased by 1.4% over the last 12 months, compared to the previous 12 month period.</b>", width = "100%")`
:::
::::

---

## 3. Sections in this release

* Medicines used in Mental Health [summary](`r config$html_top_link`)
* BNF Section 4.3 - [Antidepressant drugs](`r config$html_0403_link`)
* BNF Section 4.1 - [Hypnotics and anxiolytics](`r config$html_0401_link`)
* BNF Section 4.4 - [Central nervous system (CNS) stimulants and drugs used for ADHD](`r config$html_0404_link`)
* BNF Section 4.11 - [Drugs for dementia](`r config$html_0411_link`)
* [Prescribing during the COVID-19 pandemic](`r config$html_model_link`)
* Medicines used in Mental Health [background](`r config$background_link`)