---
title: "`r config$publication_0411_name`"
pagetitle: MUMH England quarterly summary
always_allow_html: true
output:
  word_document:
    toc: yes
    toc_depth: '3'
  html_document:
    anchor_sections: no
    css: www/style.css
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: no
---

<script>
  $(document).ready(function() {
    $('#header').prepend('<img src=`r knitr::image_uri("www/bsa_logo.svg")` alt=\"NHSBSA logo\" style=\"position:absolute; top: 0; right: 0; padding: 10px; width: 25%;\">');
  });
</script>

<style type="text/css">  body, td {
    font-size: 16px;
    font-family: sans-serif;
  }
  .list-group-item.active {
    background-color: #005EB8!important;
    border-color: #005EB8!important;
  }
    a {
        color: #005EB8!important;
    }
    .tocify {
        border-radius: 0px!important;
    }
</style>
  <html lang="en">

```{r setup, include=FALSE}
# set code chunk options to disable echo by default
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      message = FALSE)

```

<main>

# `r config$publication_sub_name` {.toc-ignore}

Published `r config$publication_date` 

`r mumhquarterly::infoBox_border(header = "Changes to these statistics", text = paste0("For this release, we have split the publication into smaller summaries,one for each BNF Section and one for the modelling of future prescribing. We are interested in any feedback about the publication or these changes, which you can send by using our ", htmltools::HTML("<a href='https://online1.snapsurveys.com/Official_Statistics_Feedback'>Official Statistics feedback survey</a>"),"."), width = "100%")`

## Summary

Between April and June 2023:

::::{.row style="display: flex; padding-bottom: 15px;"}
:::{.col-md-6}
`r mumhquarterly::infoBox_no_border("", text = "<b>1.1 million drugs for dementia items were prescribed to an estimated 250,000 identified patients.</b>", width = "100%")`
:::
:::{.col-md-6}
`r mumhquarterly::infoBox_no_border("", text = "<b>The cost of prescribed drugs for dementia items was £6.7 million.</b>", width = "100%")`
:::

:::
::::




---

## 1. Things you should know
### 1.1. Scope {.toc-ignore} 

`r mumhquarterly::infoBox_border("Background ", text = "The [NHS Long Term Plan](https://www.longtermplan.nhs.uk/) published in 2019 focuses on improving mental health care in the UK. It aims to provide more funding for mental health services, particularly for children and people with dementia. The [Core20PLUS5](https://www.england.nhs.uk/about/equality/equality-hub/core20plus5/) approach  to reduce health inequalities also seeks to improve community care for those with severe mental illnesses.", width = "100%")`

This publication provides information on medicines used to treat psychosis in England. It covers medicines prescribed in England that are then dispensed in the community in England, Scotland, Wales, Isle of Man of the Channel Islands. The publication does not include data on medicines used in hospitals, prisons or prescribed by private doctors. 

An overview including definitions can be found [here](`r config$html_top_link`) Five BNF specific summaries total are available covering medicines used in mental health, the others being for [depression](`r config$html_0403_link`),[insomnia and anxiety](`r config$html_0401_link`), [psychosis](`r config$html_0402_link`) and [attention deficit hyperactivity disorder (ADHD)](`r config$html_0404_link`). There is also a [summary](`r config$html_model_link`) exploring how the COVID-19 pandemic has affected the use of these medicines.

These summary narratives do not discuss the demographic data associated with the provision of these medicines, but this information is available in the supporting summary tables which accompany this release. Medicines are not the only way to treat mental health conditions, and this publication does not capture information on the use of other mental health services, such as psychological therapies. Therefore, this publication may not give a complete picture of the number of people receiving treatment for mental health conditions. Information on where to find related statistics can be found in Section 4.


---

## 2. Results - Drugs for dementia

`r mumhquarterly::infoBox_border(" ", text = "Dementia is a progressive clinical syndrome characterised by a range of cognitive and behavioural symptoms that can include memory loss, problems with reasoning and communication, a change in personality, and a reduced ability to carry out daily activities such as washing or dressing. Alzheimer’s disease is the most common type of dementia. Other common types of dementia include vascular dementia (where dementia is due to cerebrovascular disease), dementia with Lewy bodies (where dementia is due to protein deposits in the brain), mixed dementia, and frontotemporal dementia (where dementia is due to progressive nerve loss in either the frontal or temporal lobes of the brain).

Drugs for dementia are described in the BNF 68 section 4.11:

* BNF chemical substance 0411000D0 - Donepezil hydrochloride
* BNF chemical substance 0411000E0 - Rivastigmine
* BNF chemical substance 0411000F0 - Galantamine
* BNF chemical substance 0411000G0 - Memantine hydrochloride

More information about these medicines can be found in section 4 of this summary.", width = "100%")`

#### Figure 1: Number of prescribed items and identified patients by financial quarter - BNF 4.11: Drugs for dementia

```{r quarterly_0411, out.width="100%"}

figure_1_0411
  
```
`r get_download_button(title = "Download chart data", data = figure_1_data_0411, filename = "figure_1_0411")`

::::{.row style="display: flex; padding-bottom: 15px;"}
:::{.col-md-4}
`r mumhquarterly::infoBox_no_border("", text = "<b>1.1 million drugs for dementia items prescribed in Q1 2023/24.</b>", width = "100%")`
:::
:::{.col-md-4}
`r mumhquarterly::infoBox_no_border("", text = "<b>250,000 identified patients received a drugs for dementia item Q1 2023/24.</b>", width = "100%")`
:::
:::{.col-md-4}
`r mumhquarterly::infoBox_no_border("", text = "<b>Items for drugs for dementia declined over the initial COVID-19 period, but are returning to their pre-pandemic upward trend.</b>", width = "100%")`
:::
::::


#### Figure 2: Cost of prescribing by financial quarter - BNF 4.11: Drugs for dementia

```{r quarterly_0411_nic, out.width="100%"}

figure_2_0411
  
```
`r get_download_button(title = "Download chart data", data = figure_2_data_0411, filename = "figure_2_0411")`

::::{.row style="display: flex; padding-bottom: 15px;"}
:::{.col-md-6}
`r mumhquarterly::infoBox_no_border("", text = "<b>The cost of prescribed drugs for dementia items was £6.7 million in Q1 2023/24.</b>", width = "100%")`
:::
:::{.col-md-6}
`r mumhquarterly::infoBox_no_border("", text = "<b>The cost of drugs for dementia items have trended upwards for the last four quarters.</b>", width = "100%")`
:::
::::

#### Figure 3: Number of prescribed items and identified patients by month - BNF 4.11: Drugs for dementia

```{r monthly_0411, out.width="100%"}

figure_3_0411
```
`r get_download_button(title = "Download chart data", data = figure_3_data_0411, filename = "figure_3_0411")`

::::{.row style="display: flex; padding-bottom: 15px;"}
:::{.col-md-6}
`r mumhquarterly::infoBox_no_border("", text = "<b>On average, 200,000 patients per month received prescribing for a drugs for dementia items over the last 12 months.</b>", width = "100%")`
:::
:::{.col-md-6}
`r mumhquarterly::infoBox_no_border("", text = "<b>Drugs for dementia items increased by 4.2% over the last 12 months.</b>", width = "100%")`
:::
::::


---

## 3. Background

### 3.1. Drugs for dementia {.toc-ignore}

Dementia is a disease that leads to progressive loss of brain function typified by memory loss, confusion, speech difficulties and problems in understanding. There are over 100 different types of dementia. The most common forms are:

* Alzheimer's disease
* Vascular dementia
* Dementia with Lewy bodies
* Pick's disease
* Huntington's disease
* Alcohol-related dementia
* HIV/AIDS related dementia

Dementia mainly affects older people but can also occur in people as young as thirty due to either alcohol abuse or HIV/AIDS. No cure for dementia currently exists. However, drugs may slow the rate of decline or in some patients make a small improvement in symptoms. Despite this, disease progression is inevitable.

You can find more information about the [various types of dementia](https://www.nhs.uk/conditions/dementia/) on the NHS website.

### 3.2. Related statistics and resources {.toc-ignore}

Prescribing of medications in primary care is only part of the big picture for mental health care in England. The [NHS Digital mental health data hub](https://digital.nhs.uk/data-and-information/data-tools-and-services/data-services/mental-health-data-hub) has links to statistics on a range of related topics which may provide further context. These include a mental health bulletin, reports on children and young people's mental health surveys, dementia assessment and referral statistics, and more.

---

## 4. Further details

Further information including details about the statistics, accessibility and contact details can be found [here](`r config$html_top_link`) in the overview publication. 