---
title: "`r config$publication_0403_name`"
pagetitle: MUMH England quarterly summary
always_allow_html: true
output:
  word_document:
    toc: yes
    toc_depth: '3'
  html_document:
    anchor_sections: no
    css: www/style.css
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: no
---

<script>
  $(document).ready(function() {
    $('#header').prepend('<img src=`r knitr::image_uri("www/bsa_logo.svg")` alt=\"NHSBSA logo\" style=\"position:absolute; top: 0; right: 0; padding: 10px; width: 25%;\">');
  });
</script>

<style type="text/css">  body, td {
    font-size: 16px;
    font-family: sans-serif;
  }
  .list-group-item.active {
    background-color: #005EB8!important;
    border-color: #005EB8!important;
  }
    a {
        color: #005EB8!important;
    }
    .tocify {
        border-radius: 0px!important;
    }
</style>
  <html lang="en">

```{r setup, include=FALSE}
# set code chunk options to disable echo by default
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      message = FALSE)

```

<main>

# `r config$publication_sub_name` {.toc-ignore}

Published `r config$publication_date` 

## Summary

Between July and September 2023:

::::{.row style="display: flex; padding-bottom: 15px;"}
:::{.col-md-6}
`r infoBox_no_border("", text = "<b>22 million antidepressant items were prescribed to an estimated 6.8 million identified patients.</b>", width = "100%")`
:::
:::{.col-md-6}
`r infoBox_no_border("", text = "<b>The cost of prescribed antidepressant items was £56 million.</b>", width = "100%")`
:::
::::

---

## 1. Things you should know

`r infobox_border(" ", text = "Antidepressant drugs are effective for treating moderate to severe depression, but are not usually recommended for the routine treatment of mild depression. In those cases, psychological therapy should be considered initially. However, a trial of antidepressant therapy may be considered in cases resistant to psychological treatments or associated with other problems. The National Institute for Health and Care Excellence (NICE) has [released updated guidance](https://www.nice.org.uk/news/nice-draft-quality-standard-depression-adults-update-2023) to recommend adult patients stopping antidepressants should have their dose reduced in stages across time.

Antidepressant drugs are described in the BNF 68 section 4.3:

* BNF paragraph 4.3.1 - Tricyclic and related antidepressant drugs
* BNF paragraph 4.3.2 - Monoamine-oxidase inhibitors (MAOIs)
* BNF paragraph 4.3.3 - Selective serotonin re-uptake inhibitors (SSRIs)
* BNF paragraph 4.3.4 - Other antidepressant drugs", width = "100%")`

### 1.1. Antidepressant drugs {.toc-ignore}

Antidepressant drugs are licensed to treat major depression. Health professionals use the words depression, depressive illness or clinical depression to refer to depression. It is a serious illness and very different from the common experience of feeling unhappy or fed up for a short period of time. Depressed people may have feelings of extreme sadness that can last for a long time. These feelings are severe enough to interfere with daily life, and can last for weeks, months or years, rather than days.

It should be noted that antidepressant drugs are used for indications other than depression, for example migraine, chronic pain, myalgic encephalomyelitis (ME), or a range of other conditions. Clinical indication isn't captured by the NHSBSA. Therefore, the statistics on these drugs do not relate solely to prescribing for depression.

You can find more information about [depression](https://www.nhs.uk/conditions/clinical-depression/) on the NHS website.

---

## 2. Results - Antidepressant drugs

#### Figure 1: Number of prescribed items and identified patients by financial quarter - BNF 4.3: Antidepressant drugs

```{r quarterly_003, out.width="100%"}

figure_1_0403
  
```
`r get_download_button(title = "Download chart data", data = figure_1_data_0403, filename = "figure_1_0403")`

::::{.row style="display: flex; padding-bottom: 15px;"}
:::{.col-md-4}
`r infoBox_no_border("", text = "<b>22 million antidepressant items prescribed in Q1 2023/24.</b>", width = "100%")`
:::
:::{.col-md-4}
`r infoBox_no_border("", text = "<b>6.8 million identified patients received an antidepressant in Q2 2023/24.</b>", width = "100%")`
:::
:::{.col-md-4}
`r infoBox_no_border("", text = "<b>Antidepressants items and patients continue to show an upward trend.</b>", width = "100%")`
:::
::::


#### Figure 2: Cost of prescribing by financial quarter - BNF 4.3: Antidepressant drugs

```{r quarterly_003_nic, out.width="100%"}

figure_2_0403
  
```
`r get_download_button(title = "Download chart data", data = figure_2_data_0403, filename = "figure_2_0403")`

::::{.row style="display: flex; padding-bottom: 15px;"}
:::{.col-md-6}
`r infoBox_no_border("", text = "<b>The cost of prescribed antidepressant items decreased to £56 million in Q2 2023/24.</b>", width = "100%")`
:::
:::{.col-md-6}
`r infoBox_no_border("", text = "<b>Price fluctuations have caused a more varied trend in antidepressant costs, compared to item and patient trends.</b>", width = "100%")`
:::
::::

#### Figure 3: Number of prescribed items and identified patients by month - BNF 4.3: Antidepressant drugs

```{r monthly_0403, out.width="100%"}

figure_3_0403
```
`r get_download_button(title = "Download chart data", data = figure_3_data_0403, filename = "figure_3_0403")`

::::{.row style="display: flex; padding-bottom: 15px;"}
:::{.col-md-6}
`r infoBox_no_border("", text = "<b>An average of 4.7 million patients received an antidepressant each month over the last 12 months.</b>", width = "100%")`
:::
:::{.col-md-6}
`r infoBox_no_border("", text = "<b>Antidepressant items increased by 3.3% over the last 12 months, compared to the previous 12 month period.</b>", width = "100%")`
:::
::::


---



---

## 3. Sections in this release

* Medicines used in Mental Health [summary](`r config$html_top_link`)
* BNF Section 4.1 - [Hypnotics and anxiolytics](`r config$html_0401_link`)
* BNF Section 4.2 - [Drugs used in psychoses and related disorders](`r config$html_0402_link`)
* BNF Section 4.4 - [Central nervous system (CNS) stimulants and drugs used for ADHD](`r config$html_0404_link`)
* BNF Section 4.11 - [Drugs for dementia](`r config$html_0411_link`)
* [Prescribing during the COVID-19 pandemic](`r config$html_model_link`)
* Medicines used in Mental Health [background](`r config$background_link`)