---
title: "Medicines Used in Mental Health -  <br> Prescribing during the COVID-19 pandemic "
pagetitle: MUMH England quarterly summary
always_allow_html: true
output:
  word_document:
    toc: yes
    toc_depth: '3'
  html_document:
    anchor_sections: no
    css: www/style.css
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: no
---

<script>
  $(document).ready(function() {
    $('#header').prepend('<img src=`r knitr::image_uri("www/bsa_logo.svg")` alt=\"NHSBSA logo\" style=\"position:absolute; top: 0; right: 0; padding: 10px; width: 25%;\">');
  });
</script>

<style type="text/css">  body, td {
    font-size: 16px;
    font-family: sans-serif;
  }
  .list-group-item.active {
    background-color: #005EB8!important;
    border-color: #005EB8!important;
  }
    a {
        color: #005EB8!important;
    }
    .tocify {
        border-radius: 0px!important;
    }
</style>
  <html lang="en">

```{r setup, include=FALSE}
# set code chunk options to disable echo by default
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      message = FALSE)

```

<main>

# England April 2017 to June 2023 {.toc-ignore}

Published 14 September 2023

`r mumhquarterly::infoBox_border(header = "Changes to these statistics", text = paste0("For this release, we have split the publication into smaller summaries, one for each BNF Section and one for the modelling of future prescribing. We are interested in any feedback about the publication or these changes, which you can send by using our ", htmltools::HTML("<a href='https://online1.snapsurveys.com/Official_Statistics_Feedback'>Official Statistics feedback survey</a>"),"."), width = "100%")`

## Summary

::::{.row style="display: flex; padding-bottom: 15px;"}
:::{.col-md-6}
`r infoBox_no_border("", text = "<b> Antidepressants, hypnotics and anxiolytics, and antipsychotics have generally remained within the range of values expected based on pre-pandemic trends.</b>", width = "100%")`
:::
:::{.col-md-6}
`r infoBox_no_border("", text = "<b>CNS stimulants and drugs for ADHD have continued to increase beyond the range of expected items to be prescribed.</b>", width = "100%")`
:::
:::{.col-md-6}
`r infoBox_no_border("", text = "<b>Prescribing of drugs for dementia items has consistently been below the range of expected values for the previous 2 years.</b>", width = "100%")`
:::
::::

## 1. Prescribing during the COVID-19 pandemic 


We have created a model to help users interpret the impact of the COVID-19 pandemic on prescribing trends.

The number of expected prescription items for each BNF section during this period has been calculated by extrapolating figures for March 2020 to March 2023 according to:

* the trend in prescription items for each section between April 2015 and February 2020
* the number of dispensing days that occur in each month
* the typical differences between each month of the year observed between April 2015 and February 2020
* the typical differences between each combination of 20 year age band group and gender

A linear regression model has been used to estimate the number of items prescribed, to stay relatively simple and consistent with the model used in previous releases. The model estimates the expected number of items prescribed based on pre-pandemic trends, for each combination of age band and gender. These estimates have then been added together to get an estimate of the total items expected per month, with a 99% prediction interval (PI) calculated.

::::{.row style="display: flex; padding-bottom: 10px;"} 
:::{.col-md-12}
`r infoBox_border("Prediction interval (PI)", text = "A range calculated based on the uncertainty around the fit of the model. A 99% PI means the new observation is likely to fall within the interval of predicted values 99% of the time.", width = "100%")`
:::
::::

The data used in this model has been split into groups by patients' 20 year age band and gender. As age band and gender were included in the model, the model was only fitted to data where age band and gender were both known. This means the charts of predictions made using the model are not based on the same full data used in the rest of this publication. These charts cannot be directly compared to each other or to the quarterly and monthly charts of items prescribed. More information on the methodology behind the model can be found in the [background and methodology document](https://nhsbsa-opendata.s3.eu-west-2.amazonaws.com/mumh/MUMH_Background_Info_Methodology_v008.html).


### 1.1. Antidepressant Prescribing {.toc-ignore}

#### Figure 32: Expected and actual prescribing activity during the COVID-19 Pandemic - BNF 4.3: Antidepressant drugs

```{r covid_0403_items, out.width="100%"}

figure_32

```
`r get_download_button(title = "Download chart data", data = figure_32_data, filename = "figure_32")`

Source: [Quarterly Summary Tables - Monthly table](https://nhsbsa-opendata.s3.eu-west-2.amazonaws.com/mumh/mumh_bnf0403_Mar_23.xlsx)

Between March 2020 and June 2023 there were 275 million antidepressant items prescribed to identified patients with a known age and gender. This was only 0.49% more than the 274 million items expected based on historical trends, so antidepressant items remained well within the expected values for this period.

### 1.2. Hypnotics and anxiolytics prescribing {.toc-ignore}

#### Figure 33: Expected and actual prescribing activity during the COVID-19 Pandemic - BNF 4.1: Hypnotics and anxiolytics

```{r covid_0401_items, out.width="100%"}

figure_33

```
`r get_download_button(title = "Download chart data", data = figure_33_data, filename = "figure_33")`

Source: [Quarterly Summary Tables - Monthly table](https://nhsbsa-opendata.s3.eu-west-2.amazonaws.com/mumh/mumh_bnf0401_Mar_23.xlsx)

There were 45.2 million hypnotics and anxiolytics items prescribed between March 2020 and June 2023 to identified patients with a known age and gender. This was only 0.48% more than the 45.0 million items expected based on historical trends, so hypnotics and anxiolytics items remained well within the expected values for this period. 

### 1.3. Drugs used in psychoses and related disorders prescribing {.toc-ignore}

#### Figure 34: Expected and actual prescribing activity during the COVID-19 Pandemic - BNF 4.2: Drugs used in psychoses and related disorders

```{r covid_0402_items, out.width="100%"}

figure_34

```
`r get_download_button(title = "Download chart data", data = figure_34_data, filename = "figure_34")`

Source: [Quarterly Summary Tables - Monthly table](https://nhsbsa-opendata.s3.eu-west-2.amazonaws.com/mumh/mumh_bnf0402_Mar_23.xlsx)

There were 42.6 million antipsychotics items prescribed from March 2020 to June 2023 to identified patients with known age and gender, 1.90% less than the 43.4 million items that were expected to be prescribed. Over this period antipsychotics items stayed mostly within the range of expected values, apart from April 2020.

### 1.4. CNS stimulants and drugs used for ADHD prescribing {.toc-ignore}

#### Figure 35: Expected and actual prescribing activity during the COVID-19 Pandemic - BNF 4.4: CNS stimulants and drugs used for ADHD

```{r covid_0404_items, out.width="100%"}

figure_35

```
`r get_download_button(title = "Download chart data", data = figure_35_data, filename = "figure_35")`

Source: [Quarterly Summary Tables - Monthly table](https://nhsbsa-opendata.s3.eu-west-2.amazonaws.com/mumh/mumh_bnf0404_Mar_23.xlsx)

From March 2020 to June 2023, 6.48 million CNS stimulants and drugs used for ADHD items were prescribed to identified patients with known age and gender. This was 15.0% more than the 5.64 million items expected based on historical trends. The difference between actual and expected items prescribed has increased noticeably overall since March 2022. This upward trajectory is beyond the range of the expected values.

The model is based on data containing identified patients with known age and gender only. As CNS stimulants and drugs used for ADHD have a relatively low patient identification rate, a lower proportion of prescribing is included in the model data. For example, patient identification for this BNF section was 79.3% in Q1 of 2015/16, rising to 88.3% in Q1 of 2023/24. This may have an impact on how well the model can extrapolate expected values from the data.

### 1.5. Drugs for dementia prescribing {.toc-ignore}

#### Figure 36: Expected and actual prescribing activity during the COVID-19 Pandemic - BNF 4.11: Drugs for dementia

```{r covid_0411_items, out.width="100%"}

figure_36

```
`r get_download_button(title = "Download chart data", data = figure_36_data, filename = "figure_36")`

Source: [Quarterly Summary Tables - Monthly table](https://nhsbsa-opendata.s3.eu-west-2.amazonaws.com/mumh/mumh_bnf0411_Mar_23.xlsx)

There were 13.3 million drugs for dementia items prescribed from March 2020 to March 2023 to identified patients with known age and gender. This is significantly lower than the 15.1 million items expected to be prescribed, a difference of 11.7%. From December 2020 to June 2023, only 3 months have not fallen outside the range of expected values.


---

## 2. Background

This publication provides information on medicines used to treat mental health conditions in England. It covers medicines prescribed in England that are then dispensed in the community in England, Scotland, Wales, Isle of Man of the Channel Islands by a pharmacy, appliance contractor, dispensing doctor, or have been personally administered by a GP practice. The publication does not include data on medicines used in hospitals, prisons or prescribed by private doctors.

### 2.1. Antidepressant drugs {.toc-ignore}

Antidepressant drugs are licensed to treat major depression. Health professionals use the words depression, depressive illness or clinical depression to refer to depression. It is a serious illness and very different from the common experience of feeling unhappy or fed up for a short period of time. Depressed people may have feelings of extreme sadness that can last for a long time. These feelings are severe enough to interfere with daily life, and can last for weeks, months or years, rather than days.

It should be noted that antidepressant drugs are used for indications other than depression, for example migraine, chronic pain, myalgic encephalomyelitis (ME), or a range of other conditions. Clinical indication isn't captured by the NHSBSA. Therefore, the statistics on these drugs do not relate solely to prescribing for depression.

You can find more information about [depression](https://www.nhs.uk/conditions/clinical-depression/) on the NHS website.

### 2.2. Hypnotics and anxiolytics {.toc-ignore}

Hypnotics and anxiolytics are used to treat insomnia and anxiety respectively.

Insomnia is difficulty getting to sleep or staying asleep for long enough to feel refreshed in the morning, despite there being enough opportunity to sleep. The most common problem with insomnia is difficult falling asleep (sleep-onset insomnia). An insomniac may also experience:

* waking in the night
* not feeling refreshed after sleep and not being able to function normally during the day
* feeling irritable and tired and finding it difficult to concentrate
* waking when they have been disturbed from sleep by pain or noise
* waking early in the morning

Anxiety is a feeling of unease, such as worry or fear, which can be mild or severe. Everyone experiences feelings of anxiety at some point in their life and feeling anxious is sometimes perfectly normal. However, people with generalised anxiety disorder (GAD) find it hard to control their worries. Their feelings of anxiety are more constant and often affect their daily life. There are several conditions for which anxiety is the main symptom. Panic disorder, phobias and post-traumatic stress disorder can all cause severe anxiety.

You can find more information about [insomnia](https://www.nhs.uk/conditions/insomnia/) and [anxiety](https://www.nhs.uk/conditions/generalised-anxiety-disorder/) from the NHS website.

### 2.3. Drugs used in psychoses and related disorders {.toc-ignore}

People experiencing psychoses may report hallucinations (seeing or hearing things that other people cannot see or hear) or delusional beliefs (believing things that are not actually true), and may exhibit personality changes and thought disorder. As well as this, they may experience difficulties with social interaction and impairment in carrying out daily life activities.

You can find more information about [psychoses and related disorders](https://www.nhs.uk/conditions/psychosis/) on the NHS website.

### 2.4. Central nervous system (CNS) stimulants and drugs used for ADHD  {.toc-ignore}

Attention deficit hyperactivity disorder (ADHD) and attention deficit disorder (ADD) refer to a range of problem behaviours associated with poor attention span. These may include impulsiveness and hyperactivity, as well as inattentiveness; behaviours that often prevent children and adults from learning and socialising. ADHD is sometimes referred to as hyperkinetic disorder (HD).

You can find more information about [ADHD](https://www.nhs.uk/conditions/attention-deficit-hyperactivity-disorder-adhd/) on the NHS website.

### 2.5. Drugs for dementia {.toc-ignore}

Dementia is a disease that leads to progressive loss of brain function typified by memory loss, confusion, speech difficulties and problems in understanding. There are over 100 different types of dementia. The most common forms are:

* Alzheimer's disease
* Vascular dementia
* Dementia with Lewy bodies
* Pick's disease
* Huntington's disease
* Alcohol-related dementia
* HIV/AIDS related dementia

Dementia mainly affects older people but can also occur in people as young as thirty due to either alcohol abuse or HIV/AIDS. No cure for dementia currently exists. However, drugs may slow the rate of decline or in some patients make a small improvement in symptoms. Despite this, disease progression is inevitable.

You can find more information about the [various types of dementia](https://www.nhs.uk/conditions/dementia/) on the NHS website.

### 2.6. Related statistics and resources {.toc-ignore}

Prescribing of medications in primary care is only part of the big picture for mental health care in England. The [NHS England mental health data hub](https://digital.nhs.uk/data-and-information/data-tools-and-services/data-services/mental-health-data-hub) has links to statistics on a range of related topics which may provide further context. These include a mental health bulletin, reports on children and young people's mental health surveys, dementia assessment and referral statistics, and more.

---

## 3. About these statistics

This publication is part of a series by the NHSBSA, which can be found on our [Medicines Used in Mental Health webpage](https://www.nhsbsa.nhs.uk/statistical-collections/medicines-used-mental-health-england). This release is intended to continue the annual summary statistics released in each year and combine it with the quarterly summary for quarter 4 of the 2022/23 financial year. To learn more about how this series compares to our other publications, you can view the Official Statistics guidance table on our [statistical collections page](https://www.nhsbsa.nhs.uk/statistical-collections). 

### 3.1. Changes to this publication {.toc-ignore}

For this release, we have split the publication into smaller summaries, one for each BNF Section and one for the modelling of future prescribing.


### 3.2 Statistical Disclosure Control (SDC) {.toc-ignore}

Statistical disclosure control has been applied to these statistics. Patient count, items, and net ingredient cost (NIC) have been redacted in the supporting summary tables if they relate to fewer than 5 patients. Further information about our statistical disclosure control protocol [can be found on our website](https://www.nhsbsa.nhs.uk/policies-and-procedures).

### 3.3. Rounding {.toc-ignore}

The high-level figures in this statistical summary have been rounded where appropriate for clarity. In most cases rounding has been to three significant figures in the main text and two significant figures in the key findings and information boxes. This is to make this narrative as accessible as possible to all readers. The summary tables released with this publication allow users to investigate this data at lower levels of granularity. Figures in the supplementary tables have not been rounded.

### 3.4. Trends during COVID-19 {.toc-ignore}

The number of items for each BNF section in the 34-month period March 2020 to March 2023 were compared to the number of items expected to be prescribed based on trends from the 'pre-COVID-19' period. The monthly data for April 2015 to February 2020 was used to extrapolate expected figures for March 2020 to June 2023. These figures were extrapolated according to:

* the trends between April 2015 and February 2020
* the number of dispensing days that occur in each month
* the typical differences between each month of the year observed between April 2015 and February 2020
* the typical differences between each combination of 20 year age band group and gender

This model states expected figures if trends had continued as before but doesn't confirm causation of any differences. Factors other than COVID-19 may have influenced prescribing, such as:

*	changes in patient behaviour, demographics, or morbidity
*	external environmental factors that could drive prescribing
*	any (and all) policy initiatives and guidance
*	other local priorities and resources


### 3.5. Planned changes to this publication {.toc-ignore}

This is an experimental official statistic release. Experimental statistics are newly developed or innovative statistics. These are published so that users and stakeholders can be involved in the assessment of their suitability and quality at an early stage. We will regularly be reviewing the methodology used within the statistics.

---

## 4. Accessibility

If you need information on this website in a different format like accessible PDF, large print, easy read, audio recording or braille, you can contact us by:

**Email**: nhsbsa.accessibility@nhs.net

**Phone**: 0191 203 5318

[Find out about call charges](https://www.nhsbsa.nhs.uk/contact-us/call-charges-and-phone-numbers)

We’ll consider your request and get back to you in 5 working days.

These contact details are only for accessibility queries. This inbox is not for technical queries or IT problems. If you have a query that is not about accessibility, go to the ‘Contact us’ section of this page.

View our [Accessibility statement for Official Statistics Narratives](https://www.nhsbsa.nhs.uk/accessibility-statement-official-statistics-narratives).

---

## 5. Feedback and contact us

Feedback is important to us; we welcome any questions and comments relating to these statistics. 

You can complete a [short survey about this publication](https://online1.snapsurveys.com/interview/caf0d685-0500-48d2-8b96-1c434a956819) to help us improve the Official Statistics that we produce. All responses will remain anonymous and individuals will not be identifiable in any report that we produce.

You can view our [privacy policy](https://www.nhsbsa.nhs.uk/our-policies/privacy) on our website to see how your data is used and stored.

You can contact us by:

**Email:** statistics@nhsbsa.nhs.uk

**You can also write to us at:**

NHSBSA - Statistics  
NHS Business Services Authority  
Stella House  
Goldcrest Way  
Newburn Riverside  
Newcastle upon Tyne  
NE15 8NY

**Responsible statistician:** Kirsty Gray

</main>
