---
title: "`r config$publication_0401_name`"
pagetitle: MUMH England quarterly summary
always_allow_html: true
output:
  word_document:
    toc: yes
    toc_depth: '3'
  html_document:
    anchor_sections: no
    css: www/style.css
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: no
---

<script>
  $(document).ready(function() {
    $('#header').prepend('<img src=`r knitr::image_uri("www/bsa_logo.svg")` alt=\"NHSBSA logo\" style=\"position:absolute; top: 0; right: 0; padding: 10px; width: 25%;\">');
  });
</script>

<style type="text/css">  body, td {
    font-size: 16px;
    font-family: sans-serif;
  }
  .list-group-item.active {
    background-color: #005EB8!important;
    border-color: #005EB8!important;
  }
    a {
        color: #005EB8!important;
    }
    .tocify {
        border-radius: 0px!important;
    }
</style>
  <html lang="en">

```{r setup, include=FALSE}
# set code chunk options to disable echo by default
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      message = FALSE)

```

<main>

# `r config$publication_sub_name` {.toc-ignore}

Published `r config$publication_date` 

`r mumhquarterly::infoBox_border(header = "Changes to these statistics", text = paste0("For this release, we have split the publication into smaller summaries,one for each BNF Section and one for the modelling of future prescribing. We are interested in any feedback about the publication or these changes, which you can send by using our ", htmltools::HTML("<a href='https://online1.snapsurveys.com/Official_Statistics_Feedback'>Official Statistics feedback survey</a>"),"."), width = "100%")`

## Summary

Between April and June 2023:

::::{.row style="display: flex; padding-bottom: 15px;"}
:::{.col-md-6}
`r mumhquarterly::infoBox_no_border("", text = "<b>3.4 million hypnotics and anxiolytics items were prescribed to an estimated 1.0 million identified patients.</b>", width = "100%")`
:::
:::{.col-md-6}
`r mumhquarterly::infoBox_no_border("", text = "<b>The cost of prescribed hypnotics and anxiolytics items was £29 million.</b>", width = "100%")`
:::

:::
::::




---

## 1. Things you should know
### 1.1. Scope {.toc-ignore} 

`r mumhquarterly::infoBox_border("Background ", text = "The [NHS Long Term Plan](https://www.longtermplan.nhs.uk/) published in 2019 focuses on improving mental health care in the UK. It aims to provide more funding for mental health services, particularly for children and people with dementia. The [Core20PLUS5](https://www.england.nhs.uk/about/equality/equality-hub/core20plus5/) approach  to reduce health inequalities also seeks to improve community care for those with severe mental illnesses.", width = "100%")`

This publication provides information on medicines used to treat insomnia and anxiety in England. It covers medicines prescribed in England that are then dispensed in the community in England, Scotland, Wales, Isle of Man of the Channel Islands. The publication does not include data on medicines used in hospitals, prisons or prescribed by private doctors. 

An overview including definitions can be found [here](`r config$html_top_link`) Five BNF specific summaries total are available covering medicines used in mental health, the others being for [depression](`r config$html_0403_link`), [psychosis](`r config$html_0402_link`), [attention deficit hyperactivity disorder (ADHD)](`r config$html_0404_link`) and [dementia](`r config$html_0411_link`). There is also a [summary](`r config$html_model_link`) exploring how the COVID-19 pandemic has affected the use of these medicines.

These summary narratives do not discuss the demographic data associated with the provision of these medicines, but this information is available in the supporting summary tables which accompany this release. Medicines are not the only way to treat mental health conditions, and this publication does not capture information on the use of other mental health services, such as psychological therapies. Therefore, this publication may not give a complete picture of the number of people receiving treatment for mental health conditions. Information on where to find related statistics can be found in Section 4.


---

## 2. Results - Hypnotics and anxiolytics

`r mumhquarterly::infoBox_border(" ", text = "Hypnotics are used to treat insomnia - a difficulty getting to sleep or staying asleep long enough to feel refreshed. Most hypnotics will sedate if given during the day. Anxiolytics are used in the treatment of anxiety states and if given at night will help to induce sleep. The drugs classified as hypnotics and anxiolytics are described in the BNF 68 section 4.1:

* BNF paragraph 4.1.1 - Hypnotics
* BNF paragraph 4.1.2 - Anxiolytics
* BNF paragraph 4.1.3 - Barbiturates

More information about these medicines can be found in section 4 of this summary.", width = "100%")`

#### Figure 1: Number of prescribed items and identified patients by financial quarter - BNF 4.1: Hypnotics and anxiolytics

```{r quarterly_0401, out.width="100%"}

figure_1_0401
  
```
`r get_download_button(title = "Download chart data", data = figure_1_data_0401, filename = "figure_1_0401")`

::::{.row style="display: flex; padding-bottom: 15px;"}
:::{.col-md-4}
`r mumhquarterly::infoBox_no_border("", text = "<b>3.4 million hypnotics and anxiolytics items prescribed in Q1 2023/24.</b>", width = "100%")`
:::
:::{.col-md-4}
`r mumhquarterly::infoBox_no_border("", text = "<b>1.0 million identified patients received an hypnotics and anxiolytic item in Q1 2023/24.</b>", width = "100%")`
:::
:::{.col-md-4}
`r mumhquarterly::infoBox_no_border("", text = "<b>Items and patients continue to trend downwards for hypnotics and anxiolytics items.</b>", width = "100%")`
:::
::::


#### Figure 2: Cost of prescribing by financial quarter - BNF 4.1: Hypnotics and anxiolytics

```{r quarterly_0401_nic, out.width="100%"}

figure_2_0401
  
```
`r get_download_button(title = "Download chart data", data = figure_1_data_0401, filename = "figure_1_0401")`

::::{.row style="display: flex; padding-bottom: 15px;"}
:::{.col-md-6}
`r mumhquarterly::infoBox_no_border("", text = "<b>The cost of prescribed hypnotics and anxiolytics items was £29 million in Q1 2023/24.</b>", width = "100%")`
:::
:::{.col-md-6}
`r mumhquarterly::infoBox_no_border("", text = "<b>The long-term trend for the cost of hypnotics and anxiolytics prescribing is trending upwards but decreased in Q1 2023/24.</b>", width = "100%")`
:::
::::

#### Figure 3: Number of prescribed items and identified patients by month - BNF 4.1: Hypnotics and anxiolytics

```{r monthly_0401, out.width="100%"}

figure_3_0401
```
`r get_download_button(title = "Download chart data", data = figure_3_data_0401, filename = "figure_3_0401")`

::::{.row style="display: flex; padding-bottom: 15px;"}
:::{.col-md-6}
`r mumhquarterly::infoBox_no_border("", text = "<b>On average, 660,000 patients per month received prescribing for hypnotics and anxiolytics over the last 12 months.</b>", width = "100%")`
:::
:::{.col-md-6}
`r mumhquarterly::infoBox_no_border("", text = "<b>Hypnotics and anxiolytics items decreased by 2.4% over the last 12 months.</b>", width = "100%")`
:::
::::


---

## 3. Background

### 3.1. Hypnotics and anxiolytics {.toc-ignore}

Hypnotics and anxiolytics are used to treat insomnia and anxiety respectively.

Insomnia is difficulty getting to sleep or staying asleep for long enough to feel refreshed in the morning, despite there being enough opportunity to sleep. The most common problem with insomnia is difficult falling asleep (sleep-onset insomnia). An insomniac may also experience:

* waking in the night
* not feeling refreshed after sleep and not being able to function normally during the day
* feeling irritable and tired and finding it difficult to concentrate
* waking when they have been disturbed from sleep by pain or noise
* waking early in the morning

Anxiety is a feeling of unease, such as worry or fear, which can be mild or severe. Everyone experiences feelings of anxiety at some point in their life and feeling anxious is sometimes perfectly normal. However, people with generalised anxiety disorder (GAD) find it hard to control their worries. Their feelings of anxiety are more constant and often affect their daily life. There are several conditions for which anxiety is the main symptom. Panic disorder, phobias and post-traumatic stress disorder can all cause severe anxiety.

You can find more information about [insomnia](https://www.nhs.uk/conditions/insomnia/) and [anxiety](https://www.nhs.uk/conditions/generalised-anxiety-disorder/) from the NHS website.

### 3.2. Related statistics and resources {.toc-ignore}

Prescribing of medications in primary care is only part of the big picture for mental health care in England. The [NHS Digital mental health data hub](https://digital.nhs.uk/data-and-information/data-tools-and-services/data-services/mental-health-data-hub) has links to statistics on a range of related topics which may provide further context. These include a mental health bulletin, reports on children and young people's mental health surveys, dementia assessment and referral statistics, and more.

---

## 4. Further details

Further information including details about the statistics, accessibility and contact details can be found [here](`r config$html_top_link`) in the overview publication. 