---
title: "`r config$publication_0402_name`"
pagetitle: MUMH England quarterly summary
always_allow_html: true
output:
  word_document:
    toc: yes
    toc_depth: '3'
  html_document:
    anchor_sections: no
    css: www/style.css
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: no
---

<script>
  $(document).ready(function() {
    $('#header').prepend('<img src=`r knitr::image_uri("www/bsa_logo.svg")` alt=\"NHSBSA logo\" style=\"position:absolute; top: 0; right: 0; padding: 10px; width: 25%;\">');
  });
</script>

<style type="text/css">  body, td {
    font-size: 16px;
    font-family: sans-serif;
  }
  .list-group-item.active {
    background-color: #005EB8!important;
    border-color: #005EB8!important;
  }
    a {
        color: #005EB8!important;
    }
    .tocify {
        border-radius: 0px!important;
    }
</style>
  <html lang="en">

```{r setup, include=FALSE}
# set code chunk options to disable echo by default
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      message = FALSE)

```

<main>

# `r config$publication_sub_name` {.toc-ignore}

Published `r config$publication_date` 

`r mumhquarterly::infoBox_border(header = "Changes to these statistics", text = paste0("For this release, we have split the publication into smaller summaries,one for each BNF Section and one for the modelling of future prescribing. We are interested in any feedback about the publication or these changes, which you can send by using our ", htmltools::HTML("<a href='https://online1.snapsurveys.com/Official_Statistics_Feedback'>Official Statistics feedback survey</a>"),"."), width = "100%")`

## Summary

Between April and June 2023:

::::{.row style="display: flex; padding-bottom: 15px;"}
:::{.col-md-6}
`r mumhquarterly::infoBox_no_border("", text = "<b>3.3 million antipsychotic items were prescribed to an estimated 650,000 identified patients.</b>", width = "100%")`
:::
:::{.col-md-6}
`r mumhquarterly::infoBox_no_border("", text = "<b>The cost of prescribed antipsychotic items was £41 million.</b>", width = "100%")`
:::

:::
::::




---

## 1. Things you should know
### 1.1. Scope {.toc-ignore} 

`r mumhquarterly::infoBox_border("Background ", text = "The [NHS Long Term Plan](https://www.longtermplan.nhs.uk/) published in 2019 focuses on improving mental health care in the UK. It aims to provide more funding for mental health services, particularly for children and people with dementia. The [Core20PLUS5](https://www.england.nhs.uk/about/equality/equality-hub/core20plus5/) approach  to reduce health inequalities also seeks to improve community care for those with severe mental illnesses.", width = "100%")`

This publication provides information on medicines used to treat psychosis in England. It covers medicines prescribed in England that are then dispensed in the community in England, Scotland, Wales, Isle of Man of the Channel Islands. The publication does not include data on medicines used in hospitals, prisons or prescribed by private doctors. 

An overview including definitions can be found [here](`r config$html_top_link`) Five BNF specific summaries total are available covering medicines used in mental health, the others being for [depression](`r config$html_0403_link`),[insomnia and anxiety](`r config$html_0401_link`), , [attention deficit hyperactivity disorder (ADHD)](`r config$html_0404_link`), and [dementia](`r config$html_0411_link`). There is also a [summary](`r config$html_model_link`) exploring how the COVID-19 pandemic has affected the use of these medicines.

These summary narratives do not discuss the demographic data associated with the provision of these medicines, but this information is available in the supporting summary tables which accompany this release. Medicines are not the only way to treat mental health conditions, and this publication does not capture information on the use of other mental health services, such as psychological therapies. Therefore, this publication may not give a complete picture of the number of people receiving treatment for mental health conditions. Information on where to find related statistics can be found in Section 4.


---

## 2. Results - Drugs used in psychoses and related disorders

`r mumhquarterly::infoBox_border(" ", text = "These drugs are used to treat psychoses and related disorders, and are referred to under the umbrella term antipsychotics. These drugs are classified and described under the BNF 68 section 4.2:

* BNF paragraph 4.2.1 - Antipsychotic drugs
* BNF paragraph 4.2.2 - Antipsychotic depot injections
* BNF paragraph 4.2.3 - Drugs for mania and hypomania

More information about these medicines can be found in section 4 of this summary.", width = "100%")`

#### Figure 1: Number of prescribed items and identified patients by financial quarter - BNF 4.2: Drugs used in psychoses and related disorders

```{r quarterly_0402, out.width="100%"}

figure_1_0402
  
```
`r get_download_button(title = "Download chart data", data = figure_1_data_0402, filename = "figure_1_0402")`

::::{.row style="display: flex; padding-bottom: 15px;"}
:::{.col-md-4}
`r mumhquarterly::infoBox_no_border("", text = "<b>3.3 million antipsychotic items prescribed in Q1 2023/24.</b>", width = "100%")`
:::
:::{.col-md-4}
`r mumhquarterly::infoBox_no_border("", text = "<b>650,000 identified patients received an antipsychotic item in Q1 2023/24.</b>", width = "100%")`
:::
:::{.col-md-4}
`r mumhquarterly::infoBox_no_border("", text = "<b>Items and patients continue to trend upwards for antipsychotics.</b>", width = "100%")`
:::
::::


#### Figure 2: Cost of prescribing by financial quarter - BNF 4.2: Drugs used in psychoses and related disorders

```{r quarterly_0402_nic, out.width="100%"}

figure_2_0402
  
```
`r get_download_button(title = "Download chart data", data = figure_2_data_0402, filename = "figure_2_0402")`

::::{.row style="display: flex; padding-bottom: 15px;"}
:::{.col-md-6}
`r mumhquarterly::infoBox_no_border("", text = "<b>The cost of prescribed antipsychotic items was £41 million in Q1 2023/24.</b>", width = "100%")`
:::
:::{.col-md-6}
`r mumhquarterly::infoBox_no_border("", text = "<b>The cost of antipsychotic prescribing descreased over the last two quarters.</b>", width = "100%")`
:::
::::

#### Figure 3: Number of prescribed items and identified patients by month - BNF 4.2: Drugs used in psychoses and related disorders

```{r monthly_0402, out.width="100%"}

figure_3_0402
```
`r get_download_button(title = "Download chart data", data = figure_3_data_0402, filename = "figure_3_0402")`

::::{.row style="display: flex; padding-bottom: 15px;"}
:::{.col-md-6}
`r mumhquarterly::infoBox_no_border("", text = "<b>On average, 500,000 patients per month received prescribing for antipsychotic over the last 12 months.</b>", width = "100%")`
:::
:::{.col-md-6}
`r mumhquarterly::infoBox_no_border("", text = "<b>Antipsychotic items increased by 1.1% over the last 12 months.</b>", width = "100%")`
:::
::::


---

## 3. Background

### 3.1. Drugs used in psychoses and related disorders {.toc-ignore}

People experiencing psychoses may report hallucinations (seeing or hearing things that other people cannot see or hear) or delusional beliefs (believing things that are not actually true), and may exhibit personality changes and thought disorder. As well as this, they may experience difficulties with social interaction and impairment in carrying out daily life activities.

You can find more information about [psychoses and related disorders](https://www.nhs.uk/conditions/psychosis/) on the NHS website.

### 3.2. Related statistics and resources {.toc-ignore}

Prescribing of medications in primary care is only part of the big picture for mental health care in England. The [NHS Digital mental health data hub](https://digital.nhs.uk/data-and-information/data-tools-and-services/data-services/mental-health-data-hub) has links to statistics on a range of related topics which may provide further context. These include a mental health bulletin, reports on children and young people's mental health surveys, dementia assessment and referral statistics, and more.

---

## 4. Further details

Further information including details about the statistics, accessibility and contact details can be found [here](`r config$html_top_link`) in the overview publication. 